Pitfall of Prediction-orienterd Machine Learning Methods and Economic Value of Statistical Efficiency in On-farm Precision Experimentation

### 2022-02-21 Writing/manuscript.rmd
modified results

- Indeed, machine learning models showed better yield prediction performances than traditional regression models. On average, the out-of-sample predicted yields' RMSE of RF (`r mean_data[field_size=="37.3 ha"&model=="RF", rmse_yield] %>% round(2)` kg ha$^{-1}$) and BRF (`r mean_data[field_size=="37.3 ha"&model=="BRF", rmse_yield] %>% round(2)` kg ha$^{-1}$) models were significantly smaller than that of OLS (`r mean_data[field_size=="37.3 ha"&model=="OLS", rmse_yield] %>% round(2)` kg ha$^{-1}$) or SER (`r mean_data[field_size=="37.3 ha"&model=="SER", rmse_yield] %>% round(2)` kg ha$^{-1}$) model. Those results are consistent with existing literature (**citations**).
- However, the better yield predictions of machine learning models did not translate into better EONR predictions. The average RMSE of out-of-sample predicted EONR of RF model (`r mean_data[field_size=="37.3 ha"&model=="RF", rmse_eonr] %>% round(2)` kg ha$^{-1}$) was much larger than that of OLS (`r mean_data[field_size=="37.3 ha"&model=="OLS", rmse_eonr] %>% round(2)` kg ha$^{-1}$) or SER (`r mean_data[field_size=="37.3 ha"&model=="SER", rmse_eonr] %>% round(2)` kg ha$^{-1}$) model. BRF model's EONR prediction (RMSE `r mean_data[field_size=="37.3 ha"&model=="BRF", rmse_eonr] %>% round(2)` kg ha$^{-1}$) was better than RF model, and was similar to OLS model, but still significantly worse than SER model. Those simulation results demonstrate that despite machine learning models have better yield predictions, their EONR predictions are worse than traditional regression models.
- Ultimately, the worse EONR prediction of machine learning models resulted in lower production profit compared to the traditional regression models. The 1,000-simulation average profit of RF ($\$$`r mean_data[field_size=="37.3 ha"&model=="RF", profit] %>% round(2)` ha$^{-1}$) or BRF ($\$$`r mean_data[field_size=="37.3 ha"&model=="BRF", profit] %>% round(2)` ha$^{-1}$) model was significantly lower than that of OLS ($\$$`r mean_data[field_size=="37.3 ha"&model=="OLS", profit] %>% round(2)` ha$^{-1}$) or SER ($\$$`r mean_data[field_size=="37.3 ha"&model=="SER", profit] %>% round(2)` ha$^{-1}$) model.

## RF model is bad

- Within the machine learning models, the RF model performed especially poorly in EONR prediction. Its EONR prediction RMSE was much larger than BRF model, and its associated profit was extremely low (as shown in Figure \@ref(fig:comb-boxplot-all) panel (A)). 
- Even if we "cheat" and feed RF model with the best possible data (i.e., the true yield response parameters), its EONR profit ($\$$`r mean_data[field_size=="37.3 ha"&model=="RF_b", profit] %>% round(2)` ha$^{-1}$) was still slightly lower than BRF model ($\$$`r mean_data[field_size=="37.3 ha"&model=="BRF", profit] %>% round(2)` ha$^{-1}$) that was estimated using observed messy data. 
- It is noteworthy to mention that RF model is currently a very popular machine learning method used in many agricultural studies. Our simulation results showed that although RF may be OK for yield prediction, it is terrible in EONR prediction, and will cause big economic loss if using it in farming recommendation development. Even after spending great efforts (and financial costs) to obtain the perfect information, using RF model still won't achieve good economic performance.

@@

## Illustrative explanation

Why good yield prediction does not necessarily end up with good economic performance? That sound counterintuitive. We attempt to provide some explanations. Note this is not a statistical paper and providing the rigorous statistical proof is beyond the scope of this paper. We just try providing some illustrative explanations.

- Figure \@ref(fig:single-simu-illustrate) showed the predicted versus true yield (panel A) and EONR (panel B) of the 1,440 locations ($aunit$) from one example simulation's results (large field). The RMSE measurements associated with each model's results for this specific simulation were displayed in the figure as well. Though the machine learning models (RF, RF_b, and BRF) generated well-predicted yields such that the plotted points were well distributed along the 45 degree line, their predicted EONR were not continuous but fell into several discrete values. That results were to a large extent caused by the discrete data of five nitrogen trial rates. But that is how the on-farm experimental data generally look like in practice. On the other hand, the SER model's predicted EONR were much better aligned the 45 degree line, despite its predicted yields were not as good as the machine learning models. The OLS model's EONR predictions are also continuous, but with too big dispersion. 
- We further look into the yield response curve estimation at individual locations. Figure \@ref(fig:yield-curves) illustrated the estimated yield response curves by different models at a specific location ($aunit$). The circle point denoted the trial N rate and true (deterministic part of) yield. This figure showed that the machine learning models' predicted yields were closer to the true yield. However, the shapes of yield response curves predicted by machine learning models were much less accurate. On the other hand, SER model's predicted yield levels were heavily biased, but its predicted yield response curve had the shape much more resembling the true response curve. The EONR is determined by the yield response (mathematically, the slope of the curve), rather than the yield level. 
- Mathematically the slope of the curve: economics term, marginal effect, but the derivative measure is not commonly used in agronomy studies; Agronomy term, treatment effect, which is not the continuous derivative measure, but regarding the different N levels as discrete values. A treatment effect is defined as the change in yield from one N rate to the next level N rate. Here we have five N treatment rates, and therefore define four treatment effects. The estimated versus true treatment effects of N rates for the same simulation round's 1,440 locations were shown in Figure \@ref(fig:treatment-effect) for different models. Notice that we used the quadratic-plateau function as the true yield response in the simulation, and therefore at higher N levels the yields are more likely to approach or pass the plateau and the true treatment effect is quite small or even zero. As can be seen, the machine learning models mostly underestimated the treatment effects at all N levels. That is the major reason why machine learning models failed to predict accurate EONR. In contrast, SER model did a much better job in treatment effect predictions, particularly at the low N levels. OLS model's estimated treatment effects were actually less biased compared to machine learning models, but the variations were just very large.
- (Maybe we don't need to say this, as this is so obvious.) Caveat: The single simulation round we choose was a little cherry-picking. We chose that the round where the five models' profit results were roughly corresponding to the means as shown in Figure \@ref(fig:comb-boxplot-all). Indeed, there are lots of variations across simulation rounds, such that in come simulations machine learning models outperformed SER model. We believe the one example simulation we showed here is representative, and can make a good illustration of the yield versus EONR prediction performances.

(I feel this part is only a minor point in this study. Perhaps it's not important enough to be put in the title.)

- It is also interesting to compare the OLS and SER results. The good performance of SER may imply some good suggestion to production.
- It has been well known in the econometric literature that SER model can improve the estimation efficiency compared to OLS at the presence of unknown spatial data autocorrrelation, which is a very common in crop field characteristics data. 
- However, since it is only an efficiency issue, and SER does not correct for bias, this advantage of SER model has not received enough attention. The efficiency improvement is only regarded as a statistical property, while its economic benefit is not well recognized.
- Our simulation evidence showed that the statistical efficiency of SER model can actually translate into considerable more profitable EONR recommendations. The profit of SER model is higher than OLS model by $\$$`r (mean_data[field_size=="37.3 ha"&model=="SER", profit] - mean_data[field_size=="37.3 ha"&model=="OLS", profit]) %>% round(2)` ha$^{-1}$, $\$$`r (mean_data[field_size=="18.7 ha"&model=="SER", profit] - mean_data[field_size=="18.7 ha"&model=="OLS", profit]) %>% round(2)` ha$^{-1}$, and $\$$`r (mean_data[field_size=="9.3 ha"&model=="SER", profit] - mean_data[field_size=="9.3 ha"&model=="OLS", profit]) %>% round(2)` ha$^{-1}$ for large, medium, and small fields, respectively.
- So, using SER model to update the OLS model is likely to be economically beneficial in on-farm experimental data analysis, and should be encouraged.

>>>>

Machine learning models showed better yield prediction performances than traditional regression models. On average, the out-of-sample predicted yields' RMSE of RF (`r mean_data[field_size=="37.3 ha"&model=="RF", rmse_yield] %>% round(2)` kg ha$^{-1}$) and BRF (`r mean_data[field_size=="37.3 ha"&model=="BRF", rmse_yield] %>% round(2)` kg ha$^{-1}$) models were significantly smaller than that of OLS (`r mean_data[field_size=="37.3 ha"&model=="OLS", rmse_yield] %>% round(2)` kg ha$^{-1}$) or SER (`r mean_data[field_size=="37.3 ha"&model=="SER", rmse_yield] %>% round(2)` kg ha$^{-1}$) model. Those results are consistent with existing literature (**citations**). However, the better yield predictions from the machine learning models did not translate into better EONR predictions for them. The average RMSE of out-of-sample predicted EONR of RF model (`r mean_data[field_size=="37.3 ha"&model=="RF", rmse_eonr] %>% round(2)` kg ha$^{-1}$) was much larger than that of OLS (`r mean_data[field_size=="37.3 ha"&model=="OLS", rmse_eonr] %>% round(2)` kg ha$^{-1}$) or SER (`r mean_data[field_size=="37.3 ha"&model=="SER", rmse_eonr] %>% round(2)` kg ha$^{-1}$) model. BRF model's EONR prediction (RMSE `r mean_data[field_size=="37.3 ha"&model=="BRF", rmse_eonr] %>% round(2)` kg ha$^{-1}$) was better than RF model, and was similar to OLS model, but still significantly worse than SER model. As expected, a similar pattern of relative performance of the models are observed in terms of the profitability of site-specific EONR recomendations. The average profit of RF ($\$$`r mean_data[field_size=="37.3 ha"&model=="RF", profit] %>% round(2)` ha$^{-1}$) or BRF ($\$$`r mean_data[field_size=="37.3 ha"&model=="BRF", profit] %>% round(2)` ha$^{-1}$) model was significantly lower than that of OLS ($\$$`r mean_data[field_size=="37.3 ha"&model=="OLS", profit] %>% round(2)` ha$^{-1}$) or SER ($\$$`r mean_data[field_size=="37.3 ha"&model=="SER", profit] %>% round(2)` ha$^{-1}$) model. These results clearly demonstrate that better yield prediction does not necessarily lead to better site-specific EONR prediction and thus more profitable site-specific input recommendations. This results is a cautionary for us to not select models and methods based on their yield prediction capabilities.

It is noteworthy that RF model performed especially poor in EONR prediction because RF is currently a very popular machine learning method in yield response function estimation. Its EONR prediction RMSE was much larger than BRF model, and its associated profit was extremely low (as shown in Figure \@ref(fig:comb-boxplot-all) panel (A)). Even when the RF model is fed with the true yield response parameters ($\alpha$, $\tau$, $\eta$) directly, its EONR profit ($\$$`r mean_data[field_size=="37.3 ha"&model=="RF_b", profit] %>% round(2)` ha$^{-1}$) was still slightly lower than BRF model ($\$$`r mean_data[field_size=="37.3 ha"&model=="BRF", profit] %>% round(2)` ha$^{-1}$) that was estimated using all the observed variables. Our simulation results showed that BRF is simply a superior model than RF both in terms yield prediction and EONR predictions. Given BRF is just as easy to implement in popular statistical software like R and python, it seems hard to justify the use of RF any longer at least in the context of site-speific yield response function estimation.

Finally, while the ML methods perform worse than LM and SER on average, it does not mean they perform worse in terms of profitability in every single simualtion round. Figure \@ref() shows the percentage each model performed the best in yield prediciton (panel a), EONR prediciton (panel b), and profitability (panel c).

## Insights into the divergence in yield and EONR prediction

It may sound counterintuitive that better yield prediction does not necessarily lead to better EONR prediction and economic performance. Here, we present an illustrative example to provide some explanations for this divergence between yield and EONR predictions.

Figure \@ref(fig:single-simu-illustrate) shows the predicted versus true yield (panel A) and EONR (panel B) of the 1,440 locations ($aunit$) from a single simulation round for the large field case. The RMSE measurements associated with each model's results for this specific simulation were displayed in the figure as well. Though the machine learning models (RF, RF_b, and BRF) generated well-predicted yields such that the plotted points were well distributed along the 45 degree line, their predicted EONR were not continuous but fell into several discrete values. That results were to a large extent caused by the discrete data of five nitrogen trial rates. But that is how the on-farm experimental data generally look like in practice. On the other hand, the SER model's predicted EONR were much better aligned the 45 degree line, despite its predicted yields were not as good as the machine learning models. The OLS model's EONR predictions are also continuous, but with too big dispersion.

Figure \@ref(fig:yield-curves) illustrates the estimated yield response curves by different models at a specific location ($aunit$). The circle point denoted the trial N rate and true (deterministic part of) yield. It shows that the machine learning models' predicted yields were closer to the true yield. However, the shape of yield response curves predicted by machine learning models were much less accurate compared to that of SER. On the other hand, SER model's predicted yield "level" were heavily biased unlike the ML methods, but its predicted yield response curve resembles the true response curve much more. It is well understood and established in the agronomic community that it is not yield "level", but the slope of the yield response curve that determins EONR. Therefore, it is not surprising that the failure by the ML methods to estimate site-specific yield response curve for this location accurately resulted in poor-performing input recommendation for this location.

Severe underestimation of the impact of nitrogen on yield is not just for the single location. Figure \@ref(fig:treatment-effect) plots true and estiamted treatment effects of four treatments (<span style = "color: blue;"> Xiaofei, list cases here</span>) by model for all the locations (1440) within a single field for a single simulation round. It shows that the machine learning models severely underestimated the treatment effects at all N levels, resulting in severe underestimation of site-specific EONR.  This pattern is also observed in another simulation study by \@ref(@kakimoto2022) and also an empirical study by \@ref(Alfonso). In contrast, SER model did a much better job in treatment effect prediction, particularly at the low N levels. OLS model's estimated treatment effects were less biased compared to machine learning models, but it is much less accurate compared to SER. Due to the flexibility of the ML methods, it is easy for them to come up with wrong explanations to yield variations. As we saw earlier, the ML models explaind yield variations better (lower RMSE of yield) even though they fail to recognize the impact of nitrogen on yield. The ML models found it more beneficial to let the other variables explain the variations in yield than nitrogen.\footnote{Note that as the nitrogen level increases, the impact of nitrogen on yield declines and it is almost zero at the higher experimental nitrogen levels.} This illustrating example above is not just a outlier that we cherry-picked. Figure \@ref() shows the the density plot of the average ratio of estimated EONR to the true EONR by model for all the simulation rounds. As you can see, the ML methods tend to severely underestiamte EONR on average.

It is also interesting to compare the OLS and SER results. It is well-established that SER is more statistically efficient than simple OLS in the presence of unknown spatial data autocorrrelation, which is a very common in OFPE data. This advantage of SER does not seem to have received much attention as it is just a statistical property, economist have a tendency to focus on unbiasedness and consistency issues more than efficiency. However, in the context of site-specific EONR estiamtion, statistical efficiency means money. The more accurately estimated coefficients are, the more accurately site-specific EONRs are estimated, leading to a higher profit.

- However, since it is only an efficiency issue, and SER does not correct for bias, this advantage of SER model has not received enough attention. The efficiency improvement is only regarded as a statistical property, while its economic benefit is not well recognized.
- Our simulation evidence showed that the statistical efficiency of SER model can actually translate into considerable more profitable EONR recommendations. The profit of SER model is higher than OLS model by $\$$`r (mean_data[field_size=="37.3 ha"&model=="SER", profit] - mean_data[field_size=="37.3 ha"&model=="OLS", profit]) %>% round(2)` ha$^{-1}$, $\$$`r (mean_data[field_size=="18.7 ha"&model=="SER", profit] - mean_data[field_size=="18.7 ha"&model=="OLS", profit]) %>% round(2)` ha$^{-1}$, and $\$$`r (mean_data[field_size=="9.3 ha"&model=="SER", profit] - mean_data[field_size=="9.3 ha"&model=="OLS", profit]) %>% round(2)` ha$^{-1}$ for large, medium, and small fields, respectively.
- So, using SER model to update the OLS model is likely to be economically beneficial in on-farm experimental data analysis, and should be encouraged.

@@
